---
title: "DESeq2 + combat seq"
author: "Benji Lamp"
date: "`r Sys.Date()`"
output:
  html_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(sva)
library(tidyr)
library(here)
```

```{r}
input <- "mRNA_filtered_rawcounts.RData"
load(here("mRNA/processed_data", input))
```


# Normalization

## ComBatSeq

Strandedness as a batch effect was seen to have distorted our data when using principal component analysis as a diagnostic tool. In order to correct the batch effect, the ComBat_seq function from the sva package was used. This function adjusts for batch effects using the batch variable from our metadata and considers the experimental design (like time course variables) to prevent overfitting.

```{r}
count_matrix <- as.matrix(df)

cdf <- ComBat_seq(count_matrix,
                  batch = samples$batch,
                  group = samples$condition)
```


## DESeq2


A standard DESeq function was applied using the counts corrected by ComBat_seq and the experimental design from the 'samples' metadata frame.

```{r, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(
  countData = cdf,
  colData = samples,
  design = ~ condition
)

dds <- DESeq(dds)
res <- results(dds)
```


```{r}
# Perform variance stabilizing transformation (VST)
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)

# Extract PCA data
pcaData <- plotPCA(vsd, intgroup = "condition", returnData=TRUE)

# Add sample labels to the PCA data (optional, for reference)
pcaData$label <- rownames(pcaData)  # Use rownames as the labels

# Define your colors
myColors <- c(
  "V" = "#F8766D",
  "MP" = "#00BA38",
  "LP" = "#619CFF",
  "EL" = "#F564E3", 
  "PL" = "#FFC000" 
)


library(ggrepel)

# Reorder factor levels for 'condition' to ensure correct order in the legend
pcaData$condition <- factor(pcaData$condition, levels = c("V", "MP", "LP", "EL", "PL"))


library(ggrepel)

# Reorder factor levels for 'condition' to ensure correct order in the legend
pcaData$condition <- factor(pcaData$condition, levels = c("V", "MP", "LP", "EL", "PL"))


pca_plot <- ggplot(pcaData, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size=5) +  # Points
  stat_ellipse(aes(group = condition), linetype = "dashed", size=1.2) +  # Add trend ellipses
  geom_text_repel(aes(label = label), 
                  size = 4, 
                  max.overlaps = 5,  # Set this to limit the number of overlapping labels displayed
                  box.padding = 0.35, 
                  point.padding = 0.3, 
                  segment.color = 'grey50') +  # Repel labels to prevent overlap
  scale_color_manual(values = myColors) +
  theme_minimal() +
  labs(title = "miRNA Normalized PCA",
       x = paste0("PC1: ", round(attr(pcaData, "percentVar")[1] * 100, 1), "% variance"),
       y = paste0("PC2: ", round(attr(pcaData, "percentVar")[2] * 100, 1), "% variance"),
       color = "Condition") +  # Add a legend title for the conditions
  theme(legend.position = "right",  # Place legend on the right
        plot.title = element_text(size = 18, face = "bold"),  # Increase title size
        axis.title = element_text(size = 16),  # Increase axis title size
        axis.text = element_text(size = 14))  # Increase axis label size
pca_plot
ggsave(filename = "mirpca_plot_3x12.png", plot = pca_plot, width = 12, height = 5)
```

```{r}
# Load necessary libraries
library(DESeq2)
library(org.Bt.eg.db)
library(pheatmap)
library(AnnotationDbi)
library(RColorBrewer)

# Create the gene-cell type table
genes <- c("ITGA6", "ITGB1", "KRT5", "KRT8", "KRT14", "KRT18", "KRT19",
           "CD3E", "CD4", "CD8A", "CD8B", "GYPA", "ITGAM", "PECAM1", "PTPRC",
           "EMCN", "ENG", "S1PR1", "GAS2", "LOX", "S100A4", "THY1",
           "ADIPOQ", "CALB2", "PLIN1", "SPX", "TRARG1")

cell_types <- c(rep("Epithelium", 7), 
                rep("Immune", 8),
                rep("Endothelial", 3),
                rep("Stroma/Fibroblast", 4),
                rep("Adipocyte", 5))

gene_celltype <- data.frame(Gene = genes, CellType = cell_types, stringsAsFactors = FALSE)

# Assume vsd_matrix is already defined from your DESeq2 analysis
# vsd_matrix <- assay(vsd)

# Map Ensembl IDs to Gene Symbols
ensembl_ids_full <- rownames(vsd_matrix)
ensembl_ids_clean <- sub("\\.\\d+$", "", ensembl_ids_full)

gene_symbols <- mapIds(org.Bt.eg.db,
                       keys = ensembl_ids_clean,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")

# Create a data frame for mapping
df <- data.frame(EnsemblID_full = ensembl_ids_full,
                 EnsemblID = ensembl_ids_clean,
                 GeneSymbol = gene_symbols,
                 stringsAsFactors = FALSE)

# Subset the data frame to include only the genes from your list
df_selected <- df[df$GeneSymbol %in% gene_celltype$Gene, ]

# Get the Ensembl IDs of the selected genes
selected_ensembl_ids <- df_selected$EnsemblID_full

# Subset the vsd_matrix to include only these genes
selected_vsd_matrix <- vsd_matrix[selected_ensembl_ids, ]

# Assign gene symbols as row names
gene_symbols_selected <- df_selected$GeneSymbol
names(gene_symbols_selected) <- df_selected$EnsemblID_full
rownames(selected_vsd_matrix) <- gene_symbols_selected[rownames(selected_vsd_matrix)]

# Normalize each row (gene) to have mean = 0 and sd = 1
row_normalized_matrix <- t(scale(t(selected_vsd_matrix)))

# Annotate samples with their conditions
sample_annotations <- as.data.frame(colData(dds)[, c("condition")])

# Rename the column in sample_annotations to 'Condition'
colnames(sample_annotations) <- "Condition"

# Ensure the Condition column is a factor with the correct levels
sample_annotations$Condition <- factor(sample_annotations$Condition, 
                                       levels = c("V", "MP", "LP", "EL", "PL"))

# Create a simple annotation dataframe
annotation <- data.frame(Condition = sample_annotations$Condition)

# Ensure row names of annotation match the column names of the matrix
rownames(annotation) <- colnames(selected_vsd_matrix)

# Define colors for the condition annotations
condition_colors <- list(Condition = c(
  "V" = "#F8766D",
  "MP" = "#00BA38",
  "LP" = "#619CFF",
  "EL" = "#F564E3", 
  "PL" = "#FFC000"))

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("blue", "white", "red"))(50)

# Prepare row annotation for genes (cell types)
gene_celltype_df <- gene_celltype
rownames(gene_celltype_df) <- gene_celltype_df$Gene

# Create row annotation for cell types
annotation_row <- data.frame(CellType = gene_celltype_df[rownames(row_normalized_matrix), "CellType"])
rownames(annotation_row) <- rownames(row_normalized_matrix)

# Define colors for cell types
celltype_colors <- c(
  "Epithelium" = "lightblue",
  "Immune" = "green",
  "Endothelial" = "orange",
  "Stroma/Fibroblast" = "pink",
  "Adipocyte" = "purple"
)

# Merge annotation colors
annotation_colors <- list(
  Condition = condition_colors$Condition,
  CellType = celltype_colors
)

# Generate the heatmap with normalized data
pheatmap(row_normalized_matrix, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         annotation_col = annotation,
         annotation_row = annotation_row,
         annotation_colors = annotation_colors,
         color = heatmap_colors,
         show_rownames = TRUE, 
         show_colnames = TRUE,
         main = "Heatmap of Selected Genes with Gene Symbols (Row Normalized)")

```


```{r}
vsd_matrix <- assay(vsd)

# Select the top 50 most variable genes
top_genes <- head(order(rowVars(vsd_matrix), decreasing = TRUE), 50)

# Subset the matrix to include only these top variable genes
top_vsd_matrix <- vsd_matrix[top_genes, ]

# Normalize each row (gene) to have mean = 0 and sd = 1
row_normalized_matrix <- t(scale(t(top_vsd_matrix)))

# Annotate samples with their conditions
sample_annotations <- as.data.frame(colData(dds)[, c("condition")])

# Rename the column in sample_annotations to 'Condition'
colnames(sample_annotations) <- "Condition"

# Ensure the Condition column is a factor with the correct levels
sample_annotations$Condition <- factor(sample_annotations$Condition, 
                                       levels = c("V", "MP", "LP", "EL", "PL"))

# Create a simple annotation dataframe
annotation <- data.frame(Condition = sample_annotations$Condition)

# Ensure row names of annotation match the column names of the matrix
rownames(annotation) <- colnames(top_vsd_matrix)

# Define colors for the condition annotations
condition_colors <- list(Condition = c(
  "V" = "#F8766D",
  "MP" = "#00BA38",
  "LP" = "#619CFF",
  "EL" = "#F564E3", 
  "PL" = "#FFC000"))

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("blue", "white", "red"))(50)

# Map Ensembl IDs to gene symbols
ensembl_ids <- rownames(top_vsd_matrix)
ensembl_ids_clean <- sub("\\.\\d+$", "", ensembl_ids)

gene_symbols <- mapIds(org.Bt.eg.db,
                       keys = ensembl_ids_clean,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")

# Replace NA values with Ensembl IDs
gene_symbols[is.na(gene_symbols)] <- ensembl_ids_clean[is.na(gene_symbols)]

# Assign gene symbols as row names
rownames(top_vsd_matrix) <- gene_symbols

# Generate the heatmap with gene symbols
pheatmap(top_vsd_matrix, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         annotation_col = annotation,
         annotation_colors = condition_colors,
         color = heatmap_colors,
         show_rownames = TRUE, 
         show_colnames = TRUE,
         main = "Heatmap of Top 50 Variable Genes with Gene Symbols")


```

# normalize based on the rows, not the whole bunch. Dont cluster based on Cols

## Uncorrected 

```{r, message=FALSE, warning=FALSE}
dds_uncorected <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData = samples,
  design = ~ condition
)

dds_uncorected <- DESeq(dds_uncorected)

res_uncorected <- results(dds_uncorected)
```

```{r}
x <- ls()
x <- x[!x %in% c("res_uncorected", "dds_uncorected", "dds", "res", "normal")]
rm(list = x)
rm(x)

save.image(here("mRNA/processed_data", "mRNA_deseq_results.RData"))
```