---
title: "Enrichment Heatmap"
author: "Benji Lamp"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, warning=FALSE}
library(DESeq2)
library(sva)
library(tidyr)
library(here)
```

```{r}
input <- "mRNA_filtered_rawcounts.RData"
load(here("mRNA/processed_data", input))
```


# Normalization

## ComBatSeq

Strandedness as a batch effect was seen to have distorted our data when using principal component analysis as a diagnostic tool. In order to correct the batch effect, the ComBat_seq function from the sva package was used. This function adjusts for batch effects using the batch variable from our metadata and considers the experimental design (like time course variables) to prevent overfitting.

```{r}
count_matrix <- as.matrix(df)

cdf <- ComBat_seq(count_matrix,
                  batch = samples$batch,
                  group = samples$condition)
```


## DESeq2


A standard DESeq function was applied using the counts corrected by ComBat_seq and the experimental design from the 'samples' metadata frame.

```{r, message=FALSE, warning=FALSE}
dds <- DESeqDataSetFromMatrix(
  countData = cdf,
  colData = samples,
  design = ~ condition
)

dds <- DESeq(dds)
res <- results(dds)
```


```{r}
# 1. Create vsd_matrix
vsd_matrix <- assay(vsd)

# 2. Map Ensembl IDs → gene symbols
gene_map <- AnnotationDbi::select(
  org.Bt.eg.db,
  keys   = rownames(vsd_matrix),
  keytype = "ENSEMBL",
  columns = "SYMBOL"
)

# 3. Add gene symbols as a new column for easy lookup
vsd_df <- as.data.frame(vsd_matrix)
vsd_df$SYMBOL <- gene_map$SYMBOL[match(rownames(vsd_df), gene_map$ENSEMBL)]

# 4. Filter for your marker genes
marker_mat <- vsd_df[vsd_df$SYMBOL %in% gene_celltype$Gene, ]
rownames(marker_mat) <- marker_mat$SYMBOL
marker_mat$SYMBOL <- NULL        # remove helper column

# 5. Draw the heat‑map
ann_colors <- list(CellType = c(
  Epithelium      = "#F8766D",
  Immune          = "#00BA38",
  Endothelial     = "#619CFF",
  `Stroma/Fibroblast` = "#F564E3",
  Adipocyte       = "#FFC000"
))

pheatmap(
  marker_mat,
  scale       = "row",
  annotation_row = gene_celltype[match(rownames(marker_mat), gene_celltype$Gene), "CellType", drop = FALSE],
  annotation_colors = ann_colors,
  color       = colorRampPalette(rev(brewer.pal(9, "RdBu")))(100),
  clustering_method = "complete",
  show_rownames = TRUE,
  fontsize_row  = 8
)

```

```{r}
# Load necessary libraries
library(DESeq2)
library(org.Bt.eg.db)
library(pheatmap)
library(AnnotationDbi)
library(RColorBrewer)

# Create the gene-cell type table
genes <- c("ITGA6", "ITGB1", "KRT5", "KRT8", "KRT14", "KRT18", "KRT19",
           "CD3E", "CD4", "CD8A", "CD8B", "GYPA", "ITGAM", "PECAM1", "PTPRC",
           "EMCN", "ENG", "S1PR1", "GAS2", "LOX", "S100A4", "THY1",
           "ADIPOQ", "CALB2", "PLIN1", "SPX", "TRARG1")

cell_types <- c(rep("Epithelium", 7), 
                rep("Immune", 8),
                rep("Endothelial", 3),
                rep("Stroma/Fibroblast", 4),
                rep("Adipocyte", 5))

gene_celltype <- data.frame(Gene = genes, CellType = cell_types, stringsAsFactors = FALSE)

# Assume vsd_matrix is already defined from your DESeq2 analysis
# vsd_matrix <- assay(vsd)

# Map Ensembl IDs to Gene Symbols
ensembl_ids_full <- rownames(vsd_matrix)
ensembl_ids_clean <- sub("\\.\\d+$", "", ensembl_ids_full)

gene_symbols <- mapIds(org.Bt.eg.db,
                       keys = ensembl_ids_clean,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")

# Create a data frame for mapping
df <- data.frame(EnsemblID_full = ensembl_ids_full,
                 EnsemblID = ensembl_ids_clean,
                 GeneSymbol = gene_symbols,
                 stringsAsFactors = FALSE)

# Subset the data frame to include only the genes from your list
df_selected <- df[df$GeneSymbol %in% gene_celltype$Gene, ]

# Get the Ensembl IDs of the selected genes
selected_ensembl_ids <- df_selected$EnsemblID_full

# Subset the vsd_matrix to include only these genes
selected_vsd_matrix <- vsd_matrix[selected_ensembl_ids, ]

# Assign gene symbols as row names
gene_symbols_selected <- df_selected$GeneSymbol
names(gene_symbols_selected) <- df_selected$EnsemblID_full
rownames(selected_vsd_matrix) <- gene_symbols_selected[rownames(selected_vsd_matrix)]

# Normalize each row (gene) to have mean = 0 and sd = 1
row_normalized_matrix <- t(scale(t(selected_vsd_matrix)))

# Annotate samples with their conditions
sample_annotations <- as.data.frame(colData(dds)[, c("condition")])

# Rename the column in sample_annotations to 'Condition'
colnames(sample_annotations) <- "Condition"

# Ensure the Condition column is a factor with the correct levels
sample_annotations$Condition <- factor(sample_annotations$Condition, 
                                       levels = c("V", "MP", "LP", "EL", "PL"))

# Create a simple annotation dataframe
annotation <- data.frame(Condition = sample_annotations$Condition)

# Ensure row names of annotation match the column names of the matrix
rownames(annotation) <- colnames(selected_vsd_matrix)

# Define colors for the condition annotations
condition_colors <- list(Condition = c(
  "V" = "#F8766D",
  "MP" = "#00BA38",
  "LP" = "#619CFF",
  "EL" = "#F564E3", 
  "PL" = "#FFC000"))

# Define color palette for the heatmap
heatmap_colors <- colorRampPalette(c("blue", "white", "red"))(50)

# Prepare row annotation for genes (cell types)
gene_celltype_df <- gene_celltype
rownames(gene_celltype_df) <- gene_celltype_df$Gene

# Create row annotation for cell types
annotation_row <- data.frame(CellType = gene_celltype_df[rownames(row_normalized_matrix), "CellType"])
rownames(annotation_row) <- rownames(row_normalized_matrix)

# Define colors for cell types
celltype_colors <- c(
  "Epithelium" = "lightblue",
  "Immune" = "green",
  "Endothelial" = "orange",
  "Stroma/Fibroblast" = "pink",
  "Adipocyte" = "purple"
)

# Merge annotation colors
annotation_colors <- list(
  Condition = condition_colors$Condition,
  CellType = celltype_colors
)

## ---- everything above here stays the same ----

# Order the genes so that rows appear grouped by CellType  ▼▼  (insert here)
celltype_levels <- c("Epithelium","Immune","Endothelial",
                     "Stroma/Fibroblast","Adipocyte")

annotation_row$CellType <- factor(annotation_row$CellType,
                                  levels = celltype_levels)

row_order <- order(annotation_row$CellType)          # indices in the new order
row_normalized_matrix <- row_normalized_matrix[row_order, ]
annotation_row        <- annotation_row[row_order, , drop = FALSE]

# (optional) add blank lines between groups
gaps_row <- cumsum(table(annotation_row$CellType))[ -length(celltype_levels) ]

# ---- pheatmap call -----------------------------------------------------------
pheatmap(row_normalized_matrix,
         cluster_rows      = FALSE,
         cluster_cols      = FALSE,
         gaps_row          = gaps_row,
         annotation_col    = annotation,
         annotation_row    = annotation_row,
         annotation_colors = annotation_colors,
         color             = heatmap_colors,
         show_rownames     = TRUE,
         show_colnames     = TRUE,
         main = "Heatmap of Enriched Genes",
         filename = "celltype_heatmap.png", 
         width = 8, height = 6                # adjust size as needed
)



```

