---
title: "mrna degpatterns"
author: "Benji Lamp"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=F}
library(DESeq2)
library(here)
library(DEGreport)
library(org.Bt.eg.db)
library(AnnotationDbi)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
```

# Intro
Using deg patterns to cluster mirna based on zscore based off of a timecourse.
\ Use normalized counts from deseq

```{r}
input <- "mRNA_deseq_results.RData"
load(here("mRNA/processed_data", input))

# Load necessary libraries
library(AnnotationDbi)
library(org.Bt.eg.db)
library(openxlsx) # For exporting to Excel with row names
background <- as.data.frame(rownames(normalized_counts))
background$gene_symbol <- mapIds(org.Bt.eg.db,
                                 keys = background$`rownames(normalized_counts)`,
                                 column = "SYMBOL",
                                 keytype = "ENSEMBL",
                                 multiVals = "first")

# Reset row names to a column to retain them in the Excel file
background <- cbind("row.names" = row.names(background), background)

# Export the data frame to an Excel file
write.xlsx(background, file = "background.xlsx", rowNames = FALSE)

# Confirm success
cat("The file 'background.xlsx' has been successfully saved.\n")
```


## Metadata

This wont be used later, but its good to have i guess
```{r}
normalized_counts <- counts(dds, normalized = TRUE)

# Assuming colnames(normalized_counts) gives the column names as shown
col_names <- colnames(normalized_counts)

split_names <- strsplit(col_names, " ")

# Extract time points and cow IDs
time_points <- sapply(split_names, `[`, 1)
cow_ids <- sapply(split_names, `[`, 2)

# Create the metadata data frame
metadata <- data.frame(
  SampleID = col_names,
  Time = factor(time_points,
                levels = c("V", "MP", "LP", "EL", "PL")),
  CowID = cow_ids
)

# Set row names to match SampleID (optional)
rownames(metadata) <- metadata$SampleID
```
# padj filtering 

mRNAs that are significant in ANY pairwise comp is considered for analysis
```{r}
# List all pairwise comparisons
timepoints <- c("V", "MP", "LP", "EL", "PL")
comparisons <- combn(timepoints, 2, simplify = FALSE)

# Initialize a vector to store significant genes
significant_genes <- c()

# Loop through each comparison
for (comparison in comparisons) {
  res <- results(dds, contrast = c("condition", comparison[2], comparison[1]))
  sig_genes <- rownames(res)[which(res$padj < 0.05)]
  significant_genes <- unique(c(significant_genes, sig_genes))
}

# Subset normalized counts for significant genes only
normalized_counts <- counts(dds, normalized = TRUE)
filtered_counts <- normalized_counts[significant_genes, ]
```

# mean normalized counts table 
```{r}
colnames(filtered_counts) <- metadata$SampleID

# Calculate mean counts per time point
mean_counts <- filtered_counts %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  gather(SampleID, count, -gene) %>%
  left_join(metadata, by = "SampleID") %>%
  group_by(gene, Time) %>%
  summarize(mean_count = mean(count)) %>%
  spread(Time, mean_count)

# Convert back to matrix format
mean_counts_matrix <- as.matrix(mean_counts[, -1])
rownames(mean_counts_matrix) <- mean_counts$gene

head(mean_counts_matrix)

```

# New metadata frame 

new meta frame that only consider individual timepoint, not cow IDs. 
```{r}
# Extract the unique time points from your original metadata
time_points <- levels(metadata$Time) 
metadata_time <- data.frame(
  Time = factor(time_points, levels = c("V", "MP", "LP", "EL", "PL"))
)
rownames(metadata_time) <- time_points  

```


# Test
```{r}
library(pheatmap)

# Z-score transformation (optional, to normalize rows for better visual contrast)
z_score_matrix <- t(scale(t(filtered_counts)))  # Scale the rows (genes)

# Generate the heatmap using pheatmap
pheatmap(log2(counts(dds) + 0.5), 
         annotation_col = metadata[, c("Time", "CowID")],  # Add metadata as annotations
         cluster_rows = TRUE,  # Cluster genes (rows)
         cluster_cols = TRUE,  # Cluster samples (columns)
         show_rownames = FALSE,  # Hide gene names (optional, for large datasets)
         show_colnames = TRUE,   # Show sample names
         color = colorRampPalette(c("navy", "white", "firebrick3"))(50),  # Define color scale
         main = "Heatmap of mRNA Expression")
```

# degPatterns 

in mrna, I used a cluser cuttoff of 15 genes, \
15/845 = x/12281 \ 
x = 285
```{r}
# Run degPatterns using the mean counts
deg_patterns <- degPatterns(
  ma = mean_counts_matrix, 
  metadata = metadata_time,  
  time = "Time",
  minc = 200              
)

x <- ls()
x <- x[!x %in% c("mean_counts_matrix", "metadata_time")]
rm(list = x)
rm(x)

save.image(here("mRNA/processed_data", "degpattens_threshold_input.RData"))
```



```{r}
normalized_data <- deg_patterns$normalized

deg_plot <- degPlotCluster(
  table = normalized_data,   
  time = "Time",             
  points = F,
  cluster_column = "cluster",
  prefix_title = "Group: "   
)


deg_plot <- deg_plot + theme_minimal() + 
  labs(title = "Gene Expression Clusters Over Time") + 
  theme(legend.position = "none") # Remove the legend

# Print the plot
print(deg_plot)
```



```{r}
mrna_clusters <- deg_patterns$df

# Extract unique Ensembl IDs from your data
ensembl_ids <- unique(mrna_clusters$genes)

# Use select function from AnnotationDbi to map Ensembl IDs to Gene Symbols
mapping <- AnnotationDbi::select(
  x = org.Bt.eg.db,
  keys = ensembl_ids,
  columns = c("ENSEMBL", "SYMBOL"),
  keytype = "ENSEMBL"
)

# Merge the gene symbol mapping with your original data frame
mrna_clusters_mapped <- merge(
  mrna_clusters,
  mapping,
  by.x = "genes",
  by.y = "ENSEMBL",
  all.x = TRUE
)


mrna_deg_patterns <- deg_patterns
```
 


```{r}
x <- ls()
x <- x[!x %in% c("mrna_deg_patterns", "deg_plot","mrna_clusters_mapped", "metadata_time", "mean_counts_matrix")]
rm(list = x)
rm(x)

save.image(here("mRNA/processed_data", "mrna_deg_patterns.RData"))
```



